Determining Causes of Death with ChatGPT: A Case Study of Deaths from Verbal Autopsy in Sierra Leone from 2019-2022

### 2024-01-02 manuscript/manuscript.tex
Working on background

Reliable counts of deaths and diagnosis of their causes are crucial for public health planning and policy making --- guiding interventions and resource allocation to reduce premature deaths \cite{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, most low-income countries have reported to have either no data on deaths or have registered less than 50\% of deaths in their country, where an even fewer 8\% of these registered deaths have a Cause of Death (COD) recorded \cite{worldhealthorganizationSCOREHealthData2021}. To fill in this gap, an alternative method known as Verbal Autopsy (VA) is used to determine likely causes for deaths at scale and outside of traditional healthcare facilities \cite{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014}. VA involves trained surveyors that interview those familiar with the deceased to gather information using standardized questionnaires and open narratives, while medical professionals evaluate the gathered information to assign probable CODs \cite{worldhealthorganizationVerbalAutopsyStandards2023}.

>>>>

Reliable counts and diagnosis of deaths are crucial for public health planning and policy making, guiding intervention strategies and resource allocation that effectively saves lives and reduces premature deaths worldwide \cite{lawnMillionNeonatalDeaths2010,lassiCommunityBasedIntervention2015,liuExcessMortalityPersons2017,ewigCommunityacquiredPneumoniaEmergency2011}. However, most low-income countries have reported to have either no data on deaths or have registered less than 50\% of deaths in their country, where an even fewer 8\% of these registered deaths have a Cause of Death (COD) recorded \cite{worldhealthorganizationSCOREHealthData2021}. To fill in this gap, an alternative method known as Verbal Autopsy (VA) is used to determine likely causes for deaths at scale and outside of traditional healthcare facilities \cite{desavignyIntegratingCommunitybasedVerbal2017,thomasVerbalAutopsyHealth2018,rampatigeSystematicReviewStatistics2014}.

VA involves two major processes: survey and COD assignment \cite{worldhealthorganizationVerbalAutopsyStandards2023}. In the survey process, trained surveyors that interview those familiar with the deceased to gather information using standardized questionnaires and open narratives. In the COD assignment process, physicians evaluate the gathered information from the surveyor interviews to assign probable CODs. Although the survey process has been an effective alternative to collect mortality data at scale, the COD assignment process has been criticized to be costly and difficult to reproduce due to the reliance on physician assigned CODs \cite{mccormickProbabilisticCauseofDeathAssignment2016,morrisFactorsAssociatedPhysician2010}.
